Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-03-12 pkg/disttree/inst/draft/draft.Rnw
added citation 'Strictly Proper Scoring Rules, Prediction, and Estimation'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2049 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ca75cf4cb10efe684d275e254c35bc26ba2d388f
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2051 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9df3d18f4f2d11499fb3b49815c955313053d37a

In weather forecasting, these properties are appealing in mountainous regions
and complex terrain where small-scale effects are not well-resolved in the NWP
outputs. As these are computed at a coarser scale, there may be abrupt local
effects within a single NWP grid cell. To illustrate this in practice,
precipitation forecasts are obtained with distributional forests at 95
meteorological stations in a mountainous region in the Alps, covering mainly
Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{map}).
More specifically, a zero-censored Gaussian
distribution is employed to model 24-hour total precipitation so that the
zero-censored point mass describes the probability of no precipitation on a
given day (see Figure~\ref{axams_onestation}). Forecasts for July are
established based on data from the same month in the years 1985--2012 along with
82~covariates derived from a wide range of NWP ensemble outputs. As
Figure~\ref{axams_onestation} shows, the station-wise models yield a full
distributional forecast for each day -- here for the same day (July~24) at one
station (Axams) over four years (2009--2012) -- based on the previous 24~years
as training data. The corresponding observations conform reasonably well with
the predictions. In Section~\ref{ProbabilisticForecasting} we investigate the
performance of distributional forests in this forecasting task in more detail.
It is shown that they perform at least on par and sometimes clearly better than
three alternative zero-censored Gaussian models: a standard ensemble model
output statistics approach (EMOS, \citealp{Gneiting+Raftery+Westveld:2005}), a GAMLSS
with regressors prespecified based on meteorological expertise following
\cite{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
\citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
\citep{Messner+Mayr+Zeileis:2017} as a technique for variable selection from all
82~available regressors.

@@

In the following, it is illustrated how distributional regression forests
can solve this problem of automatically selecting relevant variables,
interactions, and potentially nonlinear effects. For fitting the forest
only the response distribution and the list of potential predictor variables
need to be specified (along with a few algorithmic details) and then
the model fit is determined by the forest in a data-driven way. Here, we
employ a zero-censored Gaussian distribution and 82~predictor variables
computed from ensemble means and spreads of various NWP outputs.
The predictive performance of the forest is compared to three other
zero-censored Gaussian models: (a) a standard basic EMOS, (b) a GAMLSS
with prespecified effects and interactions based on meteorological
knowedge/experience, and (c) a boosted GAMLSS with automatic
selection of smooth additive terms based on all 82~predictor variables.

@@

From the GEFS outputs 14 basic forecast variables are considered with up 
to 12 variations of each of these variables such as mean/maximum/minimum
or different aggregation time ranges, etc. A detailed overview is
provided in Table~\ref{covariates}, yielding 82~predictor variables in total.
\fixme{Mention that dswrf\_mean/sprd\_min is not included?}

@@

\item \emph{Distributional forest:} All 82~predictor variables are
  considered for learning a forest of 100 trees. Bootstrap
  sampling is employed for each tree using a third of the predictors
  in each split of the tree (``mtry''). Parameters are estimated by
  adaptive local likelihood based on the forest weights, as described
  in Section~\ref{Methodology}.

@@

\item \emph{Prespecified GAMLSS:} Smooth additive splines are selected for
  the most relevant predictors, based on meteorological expert knowledge
  following \cite{Stauffer+Umlauf+Messner:2017}. More specifically, based
  on the 82~available variables, 8 terms are included in the location
  submodel and 6 in the scale submodel. Both involve an interaction of
  \emph{tp} and \emph{cape} in the afternoon (between 12 and 18 UTC)
  to capture the potential for thunderstorms that frequently occur in
  summer afternoons in the Alps. The model is estimated by maximum
  penalized likelihood using a backfitting algorithm \citep{Stasinopoulos+Rigby:2007}.

\item \emph{Boosted GAMLSS:} Smooth additive splines are selected
  automatically from all 82~available variables, using non-cyclic boosting
  for parameter estimation \citep{Hofner+Mayr+Schmid:2016,Messner+Mayr+Zeileis:2017}.
  This updates the predictor terms for the location or scale submodels iteratively
  by maximizing the log-likelihood only for the variable yielding the
  highest improvement. The iteration stops early -- before fully maximizing
  the in-sample likelihood -- based on a (computationally intensive)
  out-of-bag bootstrap estimate of the log-likelihood. The grid considered for
  the number of boosting iterations (``mstop'') is: $50, 75, \dots, 975, 1000$.

@@

The predictive performance in terms of full probabilistic forecasts is
assessed using the continuous ranked probability score (CRPS, \fixme{reference}).
For each of the models this assesses the discrepancy of the
predicted distribution function $F$ from the observation $y$ by
$$
\text{CRPS}(y, F) = \int_{-\infty}^\infty (F(z) - \mathbf{1}(y \leq z))^2 dz
$$
where \(\mathbf{1}\) is the indicator function. In the subsequent
applications, the mean CRPS is always evaluated out of sample, 
either using cross-validation or a hold-out data set (2009--2012)
that was not used for learning (1985--2008). CRPS is a proper
scoring rule \fixme{citation} and lower values indicate a better fit.

@@

\subsection{Application for one station}
\label{Application for one station}
First of all we focus on one observation station, Axams in Tyrol. This 
station is selected due to its geographical closeness to Innsbruck, 
the capital of Tyrol. As for all other stations, daily observations and 
numerical weather predictions are available for the month of July of 
28~years from 1985 through the year 2012. In Figure~\ref{axams_onestation} 
the results of fitting a distributional forest to the first 24 years of the 
data set and predicting the total amount of precipitation for one specific 
day of each of the four successive years have been shown as a motivational 
example for a probabilistic forecast. Staying with this fitted model we are 
now going to analyze variable importance of the distributional forest model 
and compare its results to those of the prespecified and the boosted GAMLSS 
and the EMOS model based on residual QQ-plots.

%%%<TH> always us the term "distributional forest"
As shown in Table~\ref{model_specification} the distributional forest 
includes all of the available covariates. A single tree model would allow 
to easily analyze which covariates are used as split variables in the nodes 
and hence have a significant impact on the response variable. In a forest 
model though, for each node in each of the trees a subset of the covariates 
is chosen randomly as possible split variables to select one from. Moreover, 
each tree is built on a resampled version of the learning data. For these 
reasons the choices of split variables vary over the trees of a forest making 
it difficult to analyze the impact of covariates just by investigating the 
different trees. To get an overview of which covariates have the most influential 
effect on the response variable a standard measurement of variable importance 
for forest models has been applied. Based on the mean decrease in accuracy 
regarding the CRPS the 10 leading covariates of the above described distributional 
forest for station Axams learned on the first 24 years are listed in 
Figure~\ref{varimp}. As expected, the numerical forecast for total precipitation 
is the most influential variable. In particular, five variations of this variable 
are among the top ten of the list of covariates ordered by variable importance.

@@

After analyzing the results of the models on only one data set where 
visualization is easier, the methods are now going to be evaluated
in a cross validation framework for the station Axams. 
The data set over all 28 years is split into 7 subsets each consisting of 
4 randomly selected years. While 6 of these subsets are used as learning 
data the 7th is the testing data set for which predictions are made and 
evaluated. This is done by calculating the CRPS value for each of the 
observations in the testing data set and then averaging them. Each of the 
7 parts is selected once as testing data which leads to 7 different settings 
of which again the average values are stored. This whole procedure is repeated 
10 times. The resulting skill scores are illustrated in 
Figure~\ref{boxplots_crps_axams} using the EMOS model as the reference model 
represented by the horizontal line at height 0. This CRPS skill score boxplot 
shows that, in this setting, both GAMLSS and the distributional forest perform 
better than the EMOS model. While the two GAMLSS lead to an improvement of 
around 4 percent with the boosted version being slightly ahead, the 
distributional forest even reaches an improvement rate close to 6 percent.
<<echo=FALSE, results=hide>>=
#### cross validation rain
if(file.exists("rain_Axams_7x10.rda")){
  load("rain_Axams_7x10.rda")
} else {
  source("rain_cross_7x10.R")
  library("gamlss.cens")
  gen.cens("NO", type = "left")
  rainres <- rain_cross(stationname = "Axams", 
                        seedconst = 7, ntree = 100,
                        forest_mtry = 27,
                        gamboost_cvr = TRUE,
                        frac = FALSE)

@@

\SweaveOpts{eval = TRUE}
%\newpage
\subsection{Application for all stations}
Until now the focus has been on one observation station only. To test and 
compare the four methods under different geographical circumstances we are 
now going to apply them on all 95 stations. At each of them, the first 
24 years are used as learning data set and predictions are made for the 
ensuing 4 years of available observations. For these out-of-sample predictions 
the CRPS skill score with the EMOS model as reference model is calculated 
for each method at each station. In Figure~\ref{boxplot_crps_all} the boxplots 
of the CRPS skill score values are plotted and show that the distributional 
forest leads to the highest improvement averaged over all stations.
The green line represents the results for the station Axams.
%NAs in gamlss in 13 stations!!!\\
%for example in 13th station in complete{\_}stations\\
%\text{Error in glim.fit(f = mu.object, X = mu.X, y = y, w = w, fv = mu, os = mu.offset,  :}\\ 
%\text{NA's in the working vector or weights for parameter mu}
%\\
% sum(is.na(crps[,3]))
%<<eval = TRUE, echo = FALSE>>=
<<echo = FALSE>>=
#### prediction over all stations 24 - 4
if(file.exists("rain_pred_24to4.rda")){
  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_pred_24to4.rda")
  load("rain_pred_24to4.rda")
} else {

@@

Figure~\ref{map} shows a map of Tyrol (based on the SRTM 90m Digital Elevation 
Data, \cite{srtm}) with the location of all observation stations included in the 
data set. Each station is illustrated by a symbol with the type of symbol 
representing the method that performed best in terms of CRPS at this station. 
The color of the symbol indicates the difference between the CRPS of the 
distributional forest and the CRPS of the best performing among the other 
three models (prespecified GAMLSS, boosted GAMLSS and EMOS).
Even though for some stations, especially in East Tyrol, one of the two GAMLSS 
and for a few others the EMOS model leads to the best results, for the 
majority of stations the distributional forest performs best.

@@

From the results in Section~\ref{ProbabilisticForecasting} we can conclude 
that distributional forests provide a powerful alternative framework requiring 
only little effort to set up a model which is competitive to commonly used 
methods or even outperforms them. Even though a proper selection of covariates 
and specification of interactions in advance can lead to better fitting 
(and probably less complex) models, this presupposes profound knowledge of 
all influencing factors. Otherwise, important aspects might be missed or 
effects which do not correspond to the truth might be imposed on the model. 
This risk can be avoided by applying a boosting algorithm for an automatic 
variable selection. In this case the trade-off is often a remarkably higher 
computational effort. To build a distributional forest the user is not 
required to provide any expert knowledge or additional information about 
the setting as variable selection and detection of interactions are done 
automatically by the tree-structure of the model. This feature is 
particularly advantageous in case of a very high number of covariates. 
Therefore, the above discussed application illustrates the strengths of 
distributional forests as they offer a very flexible framework which 
enables an easy use of the method and at the same time lead to an overall 
improvement in the predictive performance in this setting. Especially 
in situations where no or only little information is available about 
possible influences, effects or interactions this novel methodology 
can be seen as a good and recommendable compromise being able to deal 
with a wide range of modelling situations in a reasonable way.

>>>>

In weather forecasting, these properties are appealing in mountainous regions
and complex terrain where small-scale effects are not well-resolved in the NWP
outputs. As these are computed at a coarser scale, there may be abrupt local
effects within a single NWP grid cell. To illustrate this in practice,
precipitation forecasts are obtained with distributional forests at 95
meteorological stations in a mountainous region in the Alps, covering mainly
Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{map}).
More specifically, a zero-censored Gaussian
distribution is employed to model 24-hour total precipitation so that the
zero-censored point mass describes the probability of no precipitation on a
given day (see Figure~\ref{axams_onestation}). Forecasts for July are
established based on data from the same month in the years 1985--2012 along with
80~covariates derived from a wide range of NWP ensemble outputs. As
Figure~\ref{axams_onestation} shows, the station-wise models yield a full
distributional forecast for each day -- here for the same day (July~24) at one
station (Axams) over four years (2009--2012) -- based on the previous 24~years
as training data. The corresponding observations conform reasonably well with
the predictions. In Section~\ref{ProbabilisticForecasting} we investigate the
performance of distributional forests in this forecasting task in more detail.
It is shown that they perform at least on par and sometimes clearly better than
three alternative zero-censored Gaussian models: a standard ensemble model
output statistics approach (EMOS, \citealp{Gneiting+Raftery+Westveld:2005}), a GAMLSS
with regressors prespecified based on meteorological expertise following
\cite{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
\citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
\citep{Messner+Mayr+Zeileis:2017} as a technique for variable selection from all
80~available regressors.

@@

In the following, it is illustrated how distributional regression forests
can solve this problem of automatically selecting relevant variables,
interactions, and potentially nonlinear effects. For fitting the forest
only the response distribution and the list of potential predictor variables
need to be specified (along with a few algorithmic details) and then
the model fit is determined by the forest in a data-driven way. Here, we
employ a zero-censored Gaussian distribution and 80~predictor variables
computed from ensemble means and spreads of various NWP outputs.
The predictive performance of the forest is compared to three other
zero-censored Gaussian models: (a) a standard basic EMOS, (b) a GAMLSS
with prespecified effects and interactions based on meteorological
knowedge/experience, and (c) a boosted GAMLSS with automatic
selection of smooth additive terms based on all 80~predictor variables.

@@

From the GEFS outputs 14 basic forecast variables are considered with up 
to 12 variations of each of these variables such as mean/maximum/minimum
or different aggregation time ranges, etc. A detailed overview is
provided in Table~\ref{covariates}, yielding 80~predictor variables in total. (Since the minimum of the ``variable downwards short wave radiation flux'' (\emph{dswrf}) over 24 hours is always 0, the ensemble mean and standard deviation of these minimal values are not considered.) \fixme{Mention that dswrf\_mean/sprd\_min is not included?}

@@

\item \emph{Distributional forest:} All 80~predictor variables are
  considered for learning a forest of 100 trees. Bootstrap
  sampling is employed for each tree using a third of the predictors
  in each split of the tree (``mtry''). Parameters are estimated by
  adaptive local likelihood based on the forest weights, as described
  in Section~\ref{Methodology}.

@@

\item \emph{Prespecified GAMLSS:} Smooth additive splines are selected for
  the most relevant predictors, based on meteorological expert knowledge
  following \cite{Stauffer+Umlauf+Messner:2017}. More specifically, based
  on the 80~available variables, 8 terms are included in the location
  submodel and 6 in the scale submodel. Both involve an interaction of
  \emph{tp} and \emph{cape} in the afternoon (between 12 and 18 UTC)
  to capture the potential for thunderstorms that frequently occur in
  summer afternoons in the Alps. The model is estimated by maximum
  penalized likelihood using a backfitting algorithm \citep{Stasinopoulos+Rigby:2007}.

\item \emph{Boosted GAMLSS:} Smooth additive splines are selected
  automatically from all 80~available variables, using non-cyclic boosting
  for parameter estimation \citep{Hofner+Mayr+Schmid:2016,Messner+Mayr+Zeileis:2017}.
  This updates the predictor terms for the location or scale submodels iteratively
  by maximizing the log-likelihood only for the variable yielding the
  highest improvement. The iteration stops early -- before fully maximizing
  the in-sample likelihood -- based on a (computationally intensive)
  out-of-bag bootstrap estimate of the log-likelihood. The grid considered for
  the number of boosting iterations (``mstop'') is: $50, 75, \dots, 975, 1000$.

@@

The predictive performance in terms of full probabilistic forecasts is
assessed using the continuous ranked probability score (CRPS, \fixme{reference}).
For each of the models this assesses the discrepancy of the
predicted distribution function $F$ from the observation $y$ by
$$
\text{CRPS}(y, F) = \int_{-\infty}^\infty (F(z) - \mathbf{1}(y \leq z))^2 dz
$$
where \(\mathbf{1}\) is the indicator function. In the subsequent
applications, the mean CRPS is always evaluated out of sample, 
either using cross-validation or a hold-out data set (2009--2012)
that was not used for learning (1985--2008). CRPS is a proper
scoring rule \citep{Gneiting+Raftery:2007} and lower values indicate a better fit.

@@

\subsection{Application for one station}
\label{Application for one station}
First of all we focus on one observation station, Axams in Tyrol. This 
station is selected due to its geographical closeness to Innsbruck, 
the capital of Tyrol. As for all other stations, daily observations and 
numerical weather predictions are available for the month of July of 
28~years from 1985 through the year 2012. In Figure~\ref{axams_onestation} 
the results of fitting a distributional forest to the first 24 years of the 
data set and predicting the total amount of precipitation for one specific 
day of each of the four successive years have been shown as a motivational 
example for a probabilistic forecast. Staying with this fitted model,
variable importance of the distributional forest model will be analyzed and its
performance will be compared to those of the prespecified and the boosted GAMLSS 
and the EMOS model based on residual QQ-plots.

%%%<TH> always us the term "distributional forest"
As shown in Table~\ref{model_specification} the distributional forest 
includes all of the available covariates. A single tree model would allow 
to easily analyze which covariates are used as split variables in the nodes 
and hence have a significant impact on the response variable. In a forest 
model though, for each node in each of the trees a subset of the covariates 
is chosen randomly as possible split variables to select one from. Moreover, 
each tree is built on a resampled version of the learning data. For these 
reasons the choices of split variables vary over the trees of a forest making 
it difficult to analyze the impact of covariates just by investigating the 
different trees. To get an overview of which covariates have the most influential 
effect on the response variable a standard measurement of variable importance 
for forest models has been applied. Based on the mean decrease in accuracy 
regarding the CRPS the 10 leading covariates of the above described distributional 
forest for station Axams learned on the first 24 years are listed in 
Figure~\ref{varimp}. As expected, the numerical forecast for total precipitation (\emph{tp}) is the most influential variable. In particular, five variations of 
this variable are among the top ten of the list of covariates ordered by 
variable importance.

@@

After analyzing the results of the models on only one data set where 
visualization is easier, the methods are now going to be evaluated
in a cross validation framework for the station Axams. 
The data set over all 28 years is split into 7 subsets each consisting of 
4 randomly selected years. While 6 of these subsets are used as learning 
data the 7th is the testing data set for which predictions are made and 
evaluated. This is done by calculating the CRPS value for each of the 
observations in the testing data set and then averaging them. Each of the 
7 parts is selected once as testing data which leads to 7 different settings 
of which again the average values are stored. This whole procedure is repeated 
10 times. The resulting skill scores are illustrated in 
Figure~\ref{boxplots_crps_axams} using the EMOS model as the reference model 
represented by the horizontal line at height 0. This CRPS skill score boxplot 
shows that, in this setting, both GAMLSS models and the distributional forest 
perform better than the EMOS model. While the two GAMLSS lead to an improvement 
of around 4 percent with the boosted version being slightly ahead, the 
distributional forest even reaches an improvement rate close to 6 percent.
<<echo=FALSE, results=hide>>=
#### cross validation rain
if(file.exists("rain_Axams_7x10.rda")){
  load("rain_Axams_7x10.rda")
} else {
  source("rain_cross_7x10.R")
  library("gamlss.cens")
  gen.cens("NO", type = "left")
  rainres <- rain_cross(stationname = "Axams", 
                        seedconst = 7, ntree = 100,
                        forest_mtry = 27,
                        gamboost_cvr = TRUE,
                        frac = FALSE)

@@

\SweaveOpts{eval = TRUE}
%\newpage
\subsection{Application for all stations}
Until now the focus has been on one observation station only. To test and 
compare the four methods under different geographical circumstances they
are applied on all 95 stations. For each station, the first 
24 years are used as learning data set and predictions are made for the 
ensuing 4 years of available observations. For these out-of-sample predictions 
the CRPS skill score with the EMOS model as reference model is calculated 
for each method at each station. In Figure~\ref{boxplot_crps_all} the boxplots 
of the CRPS skill score values are plotted illustrating that the distributional 
forest leads to the highest improvement averaged over all stations.
The green line represents the results for the station Axams.
%NAs in gamlss in 13 stations!!!\\
%for example in 13th station in complete{\_}stations\\
%\text{Error in glim.fit(f = mu.object, X = mu.X, y = y, w = w, fv = mu, os = mu.offset,  :}\\ 
%\text{NA's in the working vector or weights for parameter mu}
%\\
% sum(is.na(crps[,3]))
%<<eval = TRUE, echo = FALSE>>=
<<echo = FALSE>>=
#### prediction over all stations 24 - 4
if(file.exists("rain_pred_24to4.rda")){
  #load("~/svn/partykit/pkg/disttree/inst/draft/rain_pred_24to4.rda")
  load("rain_pred_24to4.rda")
} else {

@@

Figure~\ref{map} shows a map of Tyrol (based on the SRTM 90m Digital Elevation 
Data, \cite{srtm}) with the location of all observation stations included in the 
data set. Each station is illustrated by a symbol with the type of symbol 
representing the method that performed best in terms of CRPS at this station. 
The color of the symbol indicates the difference between the CRPS of the 
distributional forest and the CRPS of the best performing among the other 
three models (prespecified GAMLSS, boosted GAMLSS and EMOS).
Even though for some stations, especially in East Tyrol, one of the two GAMLSS
models and for a few others the EMOS model leads to the best results, for the 
majority of stations the distributional forest performs best.

@@

From the results in Section~\ref{ProbabilisticForecasting} we can conclude 
that distributional forests provide a powerful alternative framework requiring 
only little effort to set up a model which is competitive to commonly used 
methods or even outperforms them. Even though a proper selection of covariates 
and specification of interactions in advance can lead to better fitting 
(and probably less complex) models, this presupposes profound knowledge of 
all influencing factors. Otherwise, important aspects might be missed or 
effects which do not correspond to the truth might be imposed on the model. 
This risk can be avoided by applying a boosting algorithm for an automatic 
variable selection. In this case the trade-off is often a remarkably higher 
computational effort. To build a distributional forest the user is not 
required to provide any expert knowledge or additional information about 
the setting as variable selection and detection of interactions are done 
automatically by the tree-structure of the model. This feature is 
particularly advantageous in case of a very high number of covariates. 
Therefore, the above discussed application illustrates the strengths of 
distributional forests as they offer a very flexible framework which 
enables an easy use of the method and at the same time lead to an overall 
improvement in the predictive performance in this setting. Especially 
in situations where no or only little information is available about 
possible influences, effects or interactions this novel methodology 
can be seen as a good and recommendable compromise being able to deal 
with a wide range of modeling situations in a reasonable way.
